---
title: "ATACseqQC Report"
author:
    - "by GRCF"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
    html_document:
        code_folding: hide
---


### Introduction

Examine the quality of ATACseq alignment data.

### Import R libraries
```{r}
library(ATACseqQC, quietly=T)
#library(Rsamtools, quietly=T)
#library(BSgenome.Hsapiens.UCSC.hg19, quietly=T)
#library(TxDb.Hsapiens.UCSC.hg19.knownGene, quietly=T)
#library(phastCons100way.UCSC.hg19, quietly=T)
#library(ChIPpeakAnno, quietly=T)
#library(ggplot2, quietly=T)
library(GenomicAlignments, quietly=T)
library(TxDb.Mmusculus.UCSC.mm10.knownGene, quietly=T)

# location of bam file
bamfile <- snakemake@input[['bam']]
idxfile <- snakemake@input[['bai']]
idxprefix <- gsub('.bai', '', idxfile)

# infer sample id
sid <- strsplit(basename(bamfile), split='[.]')[[1]][1]
print(paste('Sample id:',sid))

# output directory
outdir <- file.path(dirname(snakemake@output[[1]]), sid)
if (! file.exists(outdir)){
    dir.create(outdir)
}

print(paste('setting TMPDIR to:', Sys.getenv('TMPDIR')))
##knitr::knit_exit()
```

### Fragment size distribution

There should be a large proportion of reads with less than 100 bp, which represents the nucleosome-free region. 
Besides, the fragment size distribution should have a clear periodicity, which is evident in the inset figure, 
indicative of nucleosome occupacy (present in integer multiples).

```{r, fig.width=7.5, fig.height=5, dpi=300}
fragSize <- fragSizeDist(bamFiles=bamfile, bamFiles.labels=sid, index=idxprefix)
```

```{r, fig.width=7.5, fig.height=5, dpi=300}
png(file.path(outdir, paste0('fragSizeDist_',sid,'_mqc','.png')), width=7.5, height=5, units='in', res=300)
fragSize <- fragSizeDist(bamFiles=bamfile, bamFiles.labels=sid, index=idxprefix)
dev.off()
```

```{r}
# load transcripts
txs <- transcripts(TxDb.Mmusculus.UCSC.mm10.knownGene)
seqlevelsStyle(txs)<- 'NCBI'
```

```{r}
# load bam file
gal <- readGAlignments(bamfile)
```

```{r}
tsse <- TSSEscore(gal, txs)

tsse

summary(tsse$values)
```

```{r fig.width=5.5, fig.height=5, dpi=300}
plot(100*(-9:10-.5), tsse$values, type="b", 
    xlab="distance to TSS",
    ylab="aggregate TSS score")
```

```{r}
png(file.path(outdir, paste0('TSS_enrichment_scores_',sid,'_mqc','.png')), width=5.5, height=5, units='in', res=300)
plot(100*(-9:10-.5), tsse$values, type="b", 
    xlab="distance to TSS",
    ylab="aggregate TSS score")
dev.off()
```

<!-- ### Nucleosome positioning -->

<!-- #### *Adjust the read start sites* -->

<!-- Tn5 transposase has been shown to bind as a dimer and inserts two adaptors into accessible DNA locations separated by 9 bp. -->

<!-- Therefore, for downstream analysis, such as peak-calling and footprinting, all reads in input bamfile need to be shifted. -->
<!-- The function shiftGAlignmentsList is used to shift the reads. By default, all reads aligning to the positive strand are offset by +4bp, -->
<!-- and all reads aligning to the negative strand are offset by -5bp. -->

<!-- The adjusted reads will be written into a new bamfile for peak calling or footprinting. -->

<!-- ```{r, fig.width=7.5, fig.height=5, dpi=300} -->
<!-- # bamfile tags to be read in -->
<!-- possibleTag <- list("integer"=c("AM", "AS", "CM", "CP", "FI", "H0", "H1", "H2", -->
<!--                                 "HI", "IH", "MQ", "NH", "NM", "OP", "PQ", "SM", -->
<!--                                 "TC", "UQ"), -->
<!--                     "character"=c("BC", "BQ", "BZ", "CB", "CC", "CO", "CQ", "CR", -->
<!--                                   "CS", "CT", "CY", "E2", "FS", "LB", "MC", "MD", -->
<!--                                   "MI", "OA", "OC", "OQ", "OX", "PG", "PT", "PU", -->
<!--                                   "Q2", "QT", "QX", "R2", "RG", "RX", "SA", "TS", -->
<!--                                   "U2")) -->

<!-- # chromosomes to consider -->
<!-- seqlev <- paste0("chr", c(1:22,"X","Y")) -->
<!-- ##seqinformation <- seqinfo(TxDb.Hsapiens.UCSC.hg19.knownGene) -->
<!-- seqinformation <- seqinfo(Hsapiens) -->

<!-- # get available tags -->
<!-- bamTop100 <- scanBam(BamFile(bamfile, yieldSize = 100), -->
<!--                     param = ScanBamParam(tag=unlist(possibleTag)))[[1]]$tag -->
<!-- tags <- names(bamTop100)[lengths(bamTop100)>0] -->
<!-- print('Tags detected in bam file:') -->
<!-- tags -->

<!-- # Adjust the read start sites -->
<!-- which <- as(seqinformation[seqlev], "GRanges") -->
<!-- # convert style chr1 --> 1 -->
<!-- seqlevelsStyle(which) <- 'NCBI' -->
<!-- gal <- readBamFile(bamfile, tag=tags, which=which, asMates=TRUE, bigFile=TRUE) -->
<!-- shiftedBamfile <- file.path(outdir, paste(sid, "shifted", "bam", sep='.')) -->
<!-- gal1 <- shiftGAlignmentsList(gal, outbam=shiftedBamfile) -->
<!-- ##saveRDS(gal1, file=file.path(outdir, paste(sid, "gal1", "rds", sep='.'))) -->
<!-- ``` -->
<!-- #### *Promoter/Transcript body (PT) score* -->

<!-- PT score is calculated as the coverage of promoter divided by the coverage of its transcript body.  -->
<!-- PT score will show if the signal is enriched in promoters. -->

<!-- ```{r, fig.width=7.5, fig.height=5, dpi=300} -->
<!-- pt <- PTscore(gal1, txs) -->
<!-- ##saveRDS(pt, file=paste(outdir,paste(sid,"PT","rds",sep='.'),sep='/')) -->

<!-- # plot -->
<!-- g <- ggplot(data.frame(log2meanCoverage=pt$log2meanCoverage, PT_Score=pt$PT_score)) -->
<!-- g <- g + geom_point(aes(x=log2meanCoverage, y=PT_Score), shape=1, size=3) -->
<!-- g <- g + theme_bw() + theme(axis.title=element_text(color='black', size=18), axis.text=element_text(color='black', size=16)) -->
<!-- g <- g + xlab("log2 mean coverage") + ylab("Promoter vs Transcript") -->
<!-- print(g) -->
<!-- ##ggsave(file.path(outdir, paste(sid,"PT.score","png",sep='.')), width=7.5, height=5, dpi=300) -->
<!-- ``` -->

<!-- #### *Nucleosome Free Regions (NFR) score* -->

<!-- NFR score is a ratio between cut signal adjacent to TSS and that flanking the corresponding TSS.  -->
<!-- Each TSS window of 400 bp is first divided into 3 sub-regions:  -->
<!-- the most upstream 150 bp (n1), the most downstream of 150 bp (n2), and the middle 100 bp (nf).  -->
<!-- Then the number of fragments with 5’ ends overlapping each region are calculated for each TSS.  -->
<!-- The NFR score for each TSS is calculated as NFR-score = log2(nf) - log2((n1+n2)/2).  -->
<!-- A plot can be generated with the NFR scores as Y-axis and the average signals of 400 bp window as X-axis,  -->
<!-- very like a MA plot for gene expression data. -->

<!-- ```{r, fig.width=7.5, fig.height=5, dpi=300} -->
<!-- nfr <- NFRscore(gal1, txs) -->
<!-- ##saveRDS(nfr, file=paste(outdir,paste(sid,"nfr","rds",sep='.'),sep='/')) -->

<!-- # plot -->
<!-- g <- ggplot(data.frame(log2meanCoverage=nfr$log2meanCoverage, NFR_Score=nfr$NFR_score)) -->
<!-- g <- g + geom_point(aes(x=log2meanCoverage, y=NFR_Score), shape=1, size=3) -->
<!-- g <- g + theme_bw() + theme(axis.title=element_text(color='black', size=18), axis.text=element_text(color='black', size=16)) -->
<!-- g <- g + theme(plot.title=element_text(color='black', hjust=0.5, size=22)) -->
<!-- g <- g + xlab("log2 mean coverage") + ylab("Nucleosome Free Regions score") -->
<!-- g <- g + ggtitle("NFRscore for 200bp flanking TSSs") -->
<!-- print(g) -->
<!-- ##ggsave(paste(outdir,paste(sid,"NFRscore","png",sep='.'),sep='/'), width=7.5, height=5, dpi=300) -->
<!-- g <- g + coord_cartesian(xlim=c(-10, 5), ylim=c(-10, 10)) -->
<!-- print(g) -->
<!-- ##ggsave(paste(outdir,paste(sid,"NFRscore","rescaled","png",sep='.'),sep='/'), width=7.5, height=5, dpi=300) -->
<!-- ``` -->

<!-- #### *Transcription Start Site (TSS) Enrichment Score* -->
<!-- TSS enrichment score is a raio between aggregate distribution of reads centered on TSSs and that flanking the corresponding TSSs. -->
<!-- TSS score = the depth of TSS (each 100bp window within 1000 bp each side) / the depth of end flanks (100bp each end). -->
<!-- TSSE score = max(mean(TSS score in each window)). -->
<!-- TSS enrichment score is calculated according to the definition at https://www.encodeproject.org/data-standards/terms/#enrichment. -->
<!-- Transcription start site (TSS) enrichment values are dependent on the reference files used; -->
<!-- cutoff values for high quality data are listed in the following table from https://www.encodeproject.org/atac-seq/. -->

<!-- For example, using hg19 Refseq TSS annotation, TSS enrichment score -->
<!-- < 6     Concerning -->
<!-- 6 - 10  Acceptable -->
<!-- > 10    Ideal -->

<!-- ```{r, fig.width=7.5, fig.height=5, dpi=300} -->
<!-- txs <- transcripts(TxDb.Hsapiens.UCSC.hg19.knownGene) -->
<!-- tsse <- TSSEscore(gal1, txs) -->
<!-- print(paste('TSS Enrichment Score =', tsse$TSSEscore)) -->
<!-- ##saveRDS(tsse, file=paste(outdir,paste(sid,"tsse","rds",sep='.'),sep='/')) -->

<!-- # plot -->
<!-- g <- ggplot(data.frame(x=100*(-9:10-.5), score=tsse$values)) -->
<!-- g <- g + geom_line(aes(x=x, y=score), color="black") -->
<!-- g <- g + geom_point(aes(x=x, y=score), shape=21, color="black", fill="white", size=3) -->
<!-- g <- g + theme_bw() + theme(axis.title=element_text(color='black', size=18), axis.text=element_text(color='black', size=16)) -->
<!-- g <- g + theme(plot.title=element_text(color='black', hjust=0.5, size=22)) -->
<!-- g <- g + xlab("distance to TSS") + ylab("aggregate TSS score") -->
<!-- g <- g + ggtitle("TSS enrichment score") -->
<!-- print(g) -->
<!-- ##ggsave(paste(outdir,paste(sid,"TSS.enrichment.score","png",sep='.'),sep='/'), width=7.5, height=5, dpi=300) -->
<!-- ``` -->
<!-- #### *Split reads* -->

<!-- The shifted reads will be split into different bins, namely nucleosome free, mononucleosome, dinucleosome, and trinucleosome. -->
<!-- Shifted reads that do not fit into any of the above bins will be discarded. -->
<!-- Splitting reads is a time-consuming step because we are using random forest to classify the fragments based on -->
<!-- fragment length, GC content and conservation scores. -->

<!-- By default, we assign the top 10% of short reads (reads below 100_bp) as nucleosome-free regions and -->
<!-- the top 10% of intermediate length reads as (reads between 180 and 247 bp) mononucleosome. -->
<!-- This serves as the training set to classify the rest of the fragments using random forest. -->
<!-- The number of the tree will be set to 2 times of square root of the length of the training set. -->

<!-- ```{r, fig.width=7.5, fig.height=5, dpi=300} -->
<!-- # define chromosomes to process -->
<!-- txs <- txs[seqnames(txs) %in% seqlev] -->
<!-- genome <- Hsapiens -->
<!-- # convert style chr1 --> 1 -->
<!-- seqlevelsStyle(txs) <- 'NCBI' -->
<!-- seqlevelsStyle(genome) <- "NCBI" -->

<!-- # split the reads into NucleosomeFree, mononucleosome, dinucleosome and trinucleosome. -->
<!-- # and save the binned alignments into bam files. -->
<!-- ##objs <- splitGAlignmentsByCut(gal1, txs=txs, genome=genome, outPath=outdir, conservation=phastCons100way.UCSC.hg19) -->
<!-- # Conservation is an optional parameter. -->
<!-- # If you do not have the conservation score or you would like to simply split the bam files using the fragment length, -->
<!-- # then you will just need to run the command without providing the conservation argument. -->
<!-- # Without setting the conservation parameter, it will run much faster. -->
<!-- objs <- splitGAlignmentsByCut(gal1, txs=txs, genome=genome, outPath=outdir) -->
<!-- ##saveRDS(objs, file=paste(outdir,paste(sid,"objs","rds",sep='.'),sep='/')) -->
<!-- ``` -->

<!-- #### *Heatmap and coverage curve for nucleosome positions* -->

<!-- By averaging the signal across all active TSSs, -->
<!-- we should observe that nucleosome-free fragments are enriched at the TSSs, -->
<!-- whereas the nucleosome-bound fragments should be enriched both upstream and downstream of the active TSSs and -->
<!-- display characteristic phasing of upstream and downstream nucleosomes. -->
<!-- Because ATAC-seq reads are concentrated at regions of open chromatin, -->
<!-- users should see a strong nucleosome signal at the +1 nucleosome, but the signal decreases at the +2, +3 and +4 nucleosomes. -->

<!-- ```{r, fig.width=5, fig.height=10, dpi=300} -->
<!-- bamfiles <- file.path(outdir, -->
<!--                     c("NucleosomeFree.bam", -->
<!--                     "mononucleosome.bam", -->
<!--                     "dinucleosome.bam", -->
<!--                     "trinucleosome.bam")) -->

<!-- TSS <- promoters(txs, upstream=0, downstream=1) -->
<!-- TSS <- unique(TSS) -->

<!-- # estimate the library size for normalization -->
<!-- librarySize <- estLibSize(bamfiles) -->

<!-- # calculate the signals around TSSs. -->
<!-- NTILE <- 101 -->
<!-- dws <- ups <- 1010 -->
<!-- sigs <- enrichedFragments(gal=objs[c("NucleosomeFree", -->
<!--                                     "mononucleosome", -->
<!--                                     "dinucleosome", -->
<!--                                     "trinucleosome")], -->
<!--                             TSS=TSS, -->
<!--                             librarySize=librarySize, -->
<!--                             seqlev=seqlev, -->
<!--                             TSS.filter=0.5, -->
<!--                             n.tile = NTILE, -->
<!--                             upstream = ups, -->
<!--                             downstream = dws) -->
<!-- ##saveRDS(sigs, file=paste(outdir,paste(sid,"sigs","rds",sep='.'),sep='/')) -->

<!-- # log2 transformed signals -->
<!-- sigs.log2 <- lapply(sigs, function(.ele) log2(.ele+1)) -->
<!-- # plot heatmap -->
<!-- ##png(paste(outdir, paste(sid,"feature","aligned","heatmap","png",sep='.'), sep='/'), width=5, height=10, units="in", res=300) -->
<!-- featureAlignedHeatmap(sigs.log2, reCenterPeaks(TSS, width=ups+dws), zeroAt=.5, n.tile=NTILE) -->
<!-- ##dev.off() -->
<!-- ``` -->

<!-- ```{r, fig.width=7.5, fig.height=5, dpi=300} -->
<!-- # get signals normalized for nucleosome-free and nucleosome-bound regions. -->
<!-- out <- featureAlignedDistribution(sigs, -->
<!--                                 reCenterPeaks(TSS, width=ups+dws), -->
<!--                                 zeroAt=.5, n.tile=NTILE, type="l", -->
<!--                                 ylab="Averaged coverage") -->

<!-- # rescale the nucleosome-free and nucleosome signals to 0~1 -->
<!-- range01 <- function(x){(x-min(x))/(max(x)-min(x))} -->
<!-- out <- apply(out, 2, range01) -->
<!-- ##png(paste(outdir, paste(sid,"feature","aligned","distribution","png",sep='.'), sep='/'), width=7.5, height=4.5, units="in", res=300) -->
<!-- matplot(out, type="l", xaxt="n", -->
<!--         xlab="Position (bp)", -->
<!--         ylab="Fraction of signal") -->
<!-- axis(1, at=seq(0, 100, by=10)+1, -->
<!--      labels=c("-1K", seq(-800, 800, by=200), "1K"), las=2) -->
<!-- abline(v=seq(0, 100, by=10)+1, lty=2, col="gray") -->
<!-- ##dev.off() -->
<!-- ``` -->

#### *collect package information*
```{r}
sessionInfo()
```
